---
title: "secuTrialR package vignette"
author: "Patrick R. Wright"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

<!-- 
add to jaml header so that the vignette is properly recognised
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
also, consider doing the html/md one so that it can open in the help pane
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

This R package provides functions for handling data from the clinical data management system (CDMS) secuTrial. The most important components are
related to loading data exports from secuTrial into R. In brief, the package aims to allow swift execution of repetitive tasks in order
to allow spending more time on the unique aspects of a dataset.

For this vignette and also to test your installation of the secuTrialR package
some example export data is delivered with the package. The data (*calcium*) was retrieved from the **lava** package and and extended
with some made up data (ID 999). Subsequently the dataset was prepared for import into a secuTrial electronic case report form (eCRF).
After import, the data was again exported and added to the secuTrialR package. The exports processed in the following are non-rectangular.

### Load the secuTrial library

```{r}
library(secuTrialR)
```

### Introduction to the dataset

To give you a brief impression of the dataset it will be briefly shown here.

```{r}
bone_mineral_dinsity_loc <- system.file("extdata", 
                                        "calcium_secuTrial.csv", 
                                        package = "secuTrialR")

bmd_data <- read.table(file = bone_mineral_dinsity_loc, sep = ";", header = TRUE)

head(bmd_data[,c("patid", "visitdate", "bmd.bmd", "bmd.grouping", "bmd.age")])
tail(bmd_data[,c("patid", "visitdate", "bmd.bmd", "bmd.grouping", "bmd.age")])
table(bmd_data$bmd.grouping)
unique(length(bmd_data$patid))
```

As you can see there are 504 unique patients grouped into "Calcium" and "Placebo". The patient with
id 999 was subsequently added to the original dataset. This record has some missing values.
These values are missing on purpose, since missing values are needed to demonstrate some downstream 
functionalities of the package. 

The central files this vignette will be using are part of the package. Since the paths end up being a little
cryptic we will store them in the following two variables. Both zip archives are secuTrial exports of the data
above.

```{r}

export_location_shortnames <- system.file("extdata",
                                         "s_export_CSV-xls_BMD.zip",
                                          package = "secuTrialR")

export_location_longnames <- system.file("extdata",
                                          "s_export_CSV-xls_longnames_BMD.zip",
                                          package = "secuTrialR")

```

### Loading an export

The code below shows how to load a secuTrial export. The output of read_secuTrial_export is a list. 

```{r}

sT_export_shortnames <- read_secuTrial_export(data_dir = export_location_shortnames)
typeof(sT_export_shortnames)
sT_export_longtnames <- read_secuTrial_export(data_dir = export_location_longnames)
typeof(sT_export_longtnames)

```

The first element of this list is another list containing the options of the export. The export options
are followed by data frames containing meta data tables and finally the central study data tables.
At this point you can also see how exports with long names and short names differ.

```{r}

names(sT_export_shortnames)
names(sT_export_longtnames)

```

The study data for the bmd secuTrial form is stored in bmd for the short table names and dem00bmd for the long table names.
As expected they are the same if compared, since the only difference is the formatting of the table names.

```{r}
head(sT_export_shortnames$bmd[,c("pat_id", "age", "grouping", "bmd")], n = 4)

all.equal(sT_export_shortnames$bmd, sT_export_longtnames$dem00bmd)

```


### Analysing data completeness

For various reasons eCRFs are frequently only partially filled. This can be commonly
observed in cohort studies of registries. Thus, before performing more thorough analyses
it is sometimes informative to get an overview of data completeness.

To asses data completeness of a secuTrial export you first need to generate the
validation overview in secuTrial and export it as an *.xlsx file. Remember to select the
"Column" and "Completion status" columns before you export the validation overview.
The validation overview for this vignette is also part of this package.

```{r}
val_ovv_location <- system.file("extdata", 
                                "bmd_validation_overview.xlsx", 
                                package = "secuTrialR")
```

In order to asses variable completeness we first require a loaded secuTrial
validation overview and a loaded secuTrial export. Both the validation
overview and the data export should be from the same time (not more than minutes
apart if possible) to ensure data integrity.

The secuTrial export has already been loaded above. The code below shows how to load the validation overview.
The validation overview contains a line for every variable with a missing or erroneous value. As explained 
earlier, patient 999 was added with missing values. As you can see the validation report only contains entries
for "Patient" 999.

```{r}
val_ovv <- read_validation_overview(data_dir = val_ovv_location)
val_ovv[,c("Patient", "Completion status", "Column")]

```

First we will asses the completeness only for "savedforms". Below you see
that "age" is missing once while "grouping" and "bmd" are both missing twice
just as reported in the validation overview.

```{r}
assess_form_variable_completeness(form = sT_export_longtnames$dem00bmd, 
                                  casenodes_table = sT_export_longtnames$casenodes,
                                  validation_overview = val_ovv, 
                                  completeness = "savedforms")
```

If you would like to know the completeness for all forms also taking forms into account
which have not been saved, then you need to set the completeness parameter to "allforms".
Note that the occ_in_vp parameter has also been set to five. This is because the patients
in the study were followed up, up to five times (i.e. five visits). Thus, if a patient 
was followed up less then five times his or her data is not complete. To set this parameter 
correctly, knowledge of the study design is necessary. As expected the completeness is 
reduced when regarding all forms.

```{r}
assess_form_variable_completeness(form = sT_export_longtnames$dem00bmd, 
                                  casenodes_table = sT_export_longtnames$casenodes,
                                  validation_overview = val_ovv,
                                  completeness = "allforms",
                                  occ_in_vp = 5)
```
